---
title: "COVID-19 updates"
author: "Marek Gierlinski"
date: '`r format(as.POSIXlt(Sys.time(), tz = "UTC"), "%d %B %Y, %H:%M:%S UTC")`'
output:
  html_document:
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
library(drake)
knitr::opts_chunk$set(
  echo = FALSE
)
```


# Data

This document presents COVID-19 data released daily by [ECDC](https://www.ecdc.europa.eu/en). Specifically, the source for this report is

[`r readd(url_covid)`](`r readd(url_covid)`).


# How data were processed

## Time alignment

The onset of the epidemic in each country is different. Many publications align data in time based on a day where number of cases/deaths crossed a certain, arbitrary level, for example 1 per million. This relies on one data point and, since data are noisy, can introduce large random shift to the entire data set. Here this isssue is curtailed by using a range of data for time alignment. This is still an arbitrary selection, but results in a reduced random error.

For this procedure a range in the response variable (case or deaths, raw counts or normalised to population) is selected based upon visual inspection of data (I know!). Then, a time shift is introduced to each country in order to bring them together. Specifically, Italy is used as a base line, and other data are shifted to minimise the root-mean-square between a given country and a best-fitting Italy loess model. The range of the response variable used is shown in cumulative plots.

## Normalisation

Most of the plots show data normalised to population (counts per million). The dashed blue lines show the rage of the response variable chosen to align countries in time.

# Cumulative data

A small selection of countries, specifically chosen to compare witht he UK.

```{r cumulative, fig.width=6, fig.height=4}
readd(fig_shifted_cases_pop)
readd(fig_shifted_deaths_pop)
```

